V4 Neural Network Model for Shape-based Feature Extraction and Object Discrimination

### 2015-05-19 article.tex
before conclusion

\subsection{IT layer}

The exact nature of the high-level visual features 
encoded by IT neurons remains to be determined.
Yet research shows that V4 neurons encode shapes 
by population responses \cite{Pasupathy2002}.
It is likely that an IT neuron collects 
the responses from a group of afferent V4 neurons 
and synthesizes a unified representation of shapes.
Thus in our model,
an IT unit combines the inputs 
from several V4 neurons into a local image feature.
The receptive fields of the connected V4 neurons 
scatter around the center of the IT receptive field.
Though there is no anatomical evidence showing that 
the spatial distribution of the connected V4 receptive fields 
is necessarily restricted to any fixed patterns,
in our model, we choose afferent V4 neurons
from a group of rectangular grid points.
Each IT neuron has 9 afferent connections from the V4 layer.
The 9 connected neurons are located in a $3\times 3$ grid.
\figurename~\ref{fig:it} demonstrates an IT receptive field 
and the receptive fields of the connected V4 neurons.
The demonstration is shown over
the output of complex cells given the image of Lena as input.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\linewidth]{images/fig3c.eps}
\caption{Demonstration of IT neuron in our model. 
IT neuron combines V4 inputs in a grid layout rotated to the strongest orientation.}
\label{fig:it}
\end{figure}

The gird is rotated to the strongest orientation 
of the central V4 input 
in order to achieve the invariance of feature orientations.
We were inspired by the SIFT \cite{Lowe1999} descriptor.
The rotation is also demonstrated in \figurename~\ref{fig:it}.
The grid is rotated to the orientation of the edge of the hat
appearing in the center of the grid.

The output of IT neurons are thus defined as a concatenation of the V4 input vectors.
\begin{equation}
\mathbf{\vec{F}}(x,y) = \mathbf{\vec{V}}(\hat{x}_1,\hat{y}_1)\|
\mathbf{\vec{V}}(\hat{x}_2,\hat{y}_2)\|
\mathbf{\vec{V}}(\hat{x}_3,\hat{y}_3)\|{}  \cdots\|
\mathbf{\vec{V}}(\hat{x}_n,\hat{y}_n),
\end{equation}
where $(\hat{x}_i,\hat{y}_i)^{\top} = 
\left(\begin{array}{ccc}
\cos\hat\theta & -\sin\hat\theta & x \\
\sin\hat\theta & \cos\hat\theta & y
\end{array}\right)\cdot(x_i,y_i,1)^{\top}$ for $i=1,2,3,\dots,n$ and 
$\hat\theta = \arg\!\max_\theta C_\theta(x,y)$.
In this equation, $\mathbf{\vec{F}}$ represents the image feature generated by IT neurons, 
$\hat\theta$ represents the strongest orientation of the central V4 input, 
$(\hat{x}_i,\hat{y}_i)$ is a group of rotated grid points
indicating the centers of receptive fields of afferent V4 neurons, 
and $(x_i,y_i)$ is a group of grid points.
In our model, $(x_i,y_i)$ is defined as the $3\times 3$ grid points fitted 
in a rectangular region.

\subsection{Object recognition}\label{sec:exp:objectrecognition}

@@

The bags of key-points approach \cite{Csurka2004} are employed.
Key-points are detected with the salient region detector 
proposed by Kadir and Brady \cite{Kadir2001}.
Features at the key-points are extracted with our model
and passed to a classifier.
We experimented with the SVM classifier
and evaluated the classifier over the Caltech 101 dataset \cite{Fei-Fei2004}.
The parameters of our model are summarized in \tablename~\ref{tab:param}.

\begin{table}
\caption{Model Parameters}
\label{tab:param}
\begin{center}
\begin{tabular}{ll}
\hline
{\bf Description} &{\bf Value}\\ 
\hline
Radius of salient region & $R$ \\
Grid size of IT unit ($|x_{i}-x_{i+1}|$) & $|x_{i}-x_{i+1}|=R$ \\
Radius of complex receptive field ($\sigma_c$) & $\sigma_c=R/3$ \\
Radius of simple receptive filed ($\sigma_s$) & $\sigma_s=\sigma_c/2$ \\
\hline
\end{tabular}
\end{center}
\end{table}

In addition to the parameters summarized in \tablename~\ref{tab:param},
the sampling density of different preferred orientations $\theta_k$
($k=1,2,3,\dots$) as in equation (\ref{equ:v4}) is also a parameter in our model.
A density of $n$ means that $\theta_k$ is taken from the range
$\{0^{\circ}, 180^{\circ}/n, 2\times 180^{\circ}/n, \dots, (n-1)\times 180^{\circ}/n\}$.
We investigate the selection of sampling density of preferred orientations.
30 training samples are randomly chosen from each category.
For each training image, 50 features are randomly chosen and passed to the classifier.
The results reported are averaged over 8 independent runs.

\begin{figure}[h]
\centering
  \subfloat[Sampling density of orientations]{
    \includegraphics[width=0.8\linewidth]{images/fig5a.eps}
    \label{fig:result:density}}\\
  \subfloat[Number of features per image]{
    \includegraphics[width=0.8\linewidth]{images/fig5b.eps}
    \label{fig:result:nfeature}}
\caption{Performance with different parameters.}
\label{fig:result}
\end{figure}

As shown in \figurename~\ref{fig:result:density},
the selection of sampling density does not exhibit obvious impact
on the performance.
With a small density, we can obtain 
both relatively low-dimensional feature vectors
and satisfactory performance.
Thus in the succeeding experiments, a sampling density of $3$ is used.

To investigate the contribution of the number of features 
on performance,
we vary the number of features chosen from each training image.
We choose 20 training samples from each category.
The results reported are also averaged over 8 independent runs.
\figurename~\ref{fig:result:nfeature} shows that 
with 50 features per image the result approaches the best performance.
The number of required features can be notably reduced by introducing 
a clustering step in the bag of words approach \cite{Csurka2004}.
However, Serre et al. \cite{Serre2007} reported that clustering could
be sensitive to the choice of parameters and initializations,
leading poorer results.

The influence of the number of training samples 
on the performance of our model is also studied.
For each category, we randomly select training sets of size
5,10,15,20,25 and 30.
\figurename~\ref{fig:result:ntrain} shows the performance as
a function of the number of training images.
The results reported are averaged over all categories
with 50 features extracted from each training image.
The performance is compared with other published results
\cite{Bosch2007,Bosch2007b,Zhang2007,Lazebnik2006,Wang2006,Grauman2005}.

\begin{figure}[h]
\centering
  \subfloat[Comparison with other methods]{
    \includegraphics[width=0.8\linewidth]{images/fig5c.eps}
    \label{fig:result:ntrain}}\\
  \subfloat[Comparison with other features]{
    \includegraphics[width=0.8\linewidth]{images/fig5d.eps}
    \label{fig:result:compare}}
\caption{Comparison of performance.}
\label{fig:result2}
\end{figure}

\figurename~\ref{fig:result:compare} shows
the performance of our model over the 101 categories
compared with SIFT features \cite{Lowe1999},
PHOW features \cite{Lazebnik2006},
and biologically inspired HMAX model \cite{Serre2007}.
The results are obtained with 30 training images randomly selected 
from each category.
SIFT features and neural features generated with our model
are passed to an SVM classifier.
The images passed to the HMAX model are resized 
to a uniform height of 140 pixels as recommended by \cite{Serre2007}.
Our model exhibits competitive performance 
compared with other models.
As shown in \figurename~\ref{fig:result:compare},
the SIFT performance points over the 101 categories
are almost completely to the left of the equivalent line.
The PHOW performance points spread through 
a wider range from $0.5$ to $1$ indicating 
the lack of robustness compared with our model.
Compared with the biologically inspired HMAX model,
our performance is also better for more than half
of the 101 categories.

>>>>

\subsection{Object recognition layer}

As we have reviewed in section 2,
the IT cortex is where object recognition is finally achieved in the brain.
The exact nature of the high-level visual features 
encoded by IT neurons remains to be determined.
Yet research shows that V4 neurons encode shapes by population responses \cite{pasupathy2002}.
It is likely that an IT neuron collects the responses from a group of afferent V4 neurons 
and synthesizes a unified representation of shapes.
In our model, the object recognition layer consists of IT computation units.
Each IT unit combines the output of several V4 units into a local image feature.
The receptive fields of these V4 units scatter around the center of the IT receptive field.
Though there is no anatomical evidence showing that 
the spatial distribution of the V4 receptive fields 
is necessarily restricted to any fixed patterns,
in our model, we choose afferent V4 units from a group of rectangular grid points.
Each IT unit has 9 afferent connections from the V4 layer.
The 9 connected V4 units are located in a $3\times 3$ grid.
\figurename~\ref{fig:13} demonstrates an IT receptive field 
and the receptive fields of the connected V4 neurons.
The demonstration is shown over
the output of complex cells given the image of Lena as input.

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{images/fig-13.pdf}
\caption{Demonstration of IT unit in our model. 
IT unit combines V4 inputs in a grid layout rotated to the strongest orientation.}
\label{fig:13}
\end{figure}

In order to achieve the invariance of feature orientations,
the gird is rotated to the strongest orientation at the center of the receptive field.
We have been inspired by the SIFT \cite{lowe1999} descriptor.
The rotation is also demonstrated in \figurename~\ref{fig:13}.
The grid is rotated to the orientation of the edge of the hat appearing at the center of the grid.
The strongest central orientation can be obtained by taking a maximal value 
over the output of the complex cells connected to the central V4 units.

The output of an IT unit is thus defined as a concatenation of the V4 feature vectors.
\begin{equation}
\mathbf{F}(x,y) = \mathbf{V}(\hat{x}_1,\hat{y}_1)\|
\mathbf{V}(\hat{x}_2,\hat{y}_2)\|
\mathbf{V}(\hat{x}_3,\hat{y}_3)\|{}  \cdots\|
\mathbf{V}(\hat{x}_n,\hat{y}_n),
\end{equation}
where $(\hat{x}_i,\hat{y}_i)^{\top} = 
\left(\begin{array}{ccc}
\cos\hat\theta & -\sin\hat\theta & x \\
\sin\hat\theta & \cos\hat\theta & y
\end{array}\right)\cdot(x_i,y_i,1)^{\top}$ for $i=1,2,3,\dots,n$ and 
$\hat\theta = \arg\!\max_\theta C_\theta(x,y)$.
In this equation, $\mathbf{F}$ represents the image feature generated by IT units.
$\hat\theta$ represents the strongest orientation at the center of the receptive field,
which is calculated over the output of complex cells $C_\theta$.
$(\hat{x}_i,\hat{y}_i)$ is a group of rotated grid points
indicating the centers of receptive fields of afferent V4 units.
$(x_i,y_i)$ is a group of grid points.
In our model, $(x_i,y_i)$ is defined as the $3\times 3$ grid points fitted 
in a rectangular region.

\subsection{Object recognition}

@@

The bags of key-points approach \cite{csurka2004} are employed.
Key-points are detected with the saliency sublayer of the V4 computation units in our model.
Features at the key-points are extracted with IT units and passed to a classifier.
We experimented with the SVM classifier
and evaluated the classifier over the Caltech 101 dataset \cite{feifei2004}.

The parameters that controls the size of simple, complex and V4 receptive field
are summarized in \tablename~\ref{tab:1} in section 3.
In addition to these parameters,
the sampling density of different preferred orientations is also a parameter in our model.
In the RBM encoder sublayer of V4 units,
the visible RBM units collect their input from complex cells 
with different preferred orientations.
A sampling density of $n$ means that a preferred orientation is taken from the range
$\{0^{\circ}, 180^{\circ}/n, 2\times 180^{\circ}/n, \dots, (n-1)\times 180^{\circ}/n\}$.
We investigate the selection of sampling density of preferred orientations.
30 training samples are randomly chosen from each category.
For each training image, 50 features are randomly chosen and passed to the classifier.
The results reported are averaged over 8 independent runs.

\begin{figure}
\centering
  \subfloat[Sampling density of orientations]{
    \includegraphics[width=0.9\linewidth]{images/fig-14a.pdf}
    \label{fig:14a}}\\
  \subfloat[Number of features per image]{
    \includegraphics[width=0.9\linewidth]{images/fig-14b.pdf}
    \label{fig:14b}}
\caption{Performance with different parameters.}
\label{fig:14}
\end{figure}

As shown in \figurename~\ref{fig:14a},
the selection of sampling density does not exhibit obvious impact on the performance.
With a small density, we can lower the dimension of the input of RBM encoders
without a significant loss in object recognition performance.
Thus in the succeeding experiments, a sampling density of $3$ is used.

We can also manipulate the saliency threshold in the saliency sublayer of V4 units
to change the number of key-points detected in an image,
which inturn changes the number of features generated by IT units.
To investigate the contribution of the number of features on performance,
we vary the number of features chosen from each training image.
We choose 20 training samples from each category.
The results reported are also averaged over 8 independent runs.
\figurename~\ref{fig:14b} shows that 
with 50 features per image the result approaches the best performance.
The number of required features can be notably reduced by introducing 
a clustering step in the bag of words approach \cite{csurka2004}.
However, Serre et al. \cite{serre2007} reported that clustering could
be sensitive to the choice of parameters and initializations,
leading poorer results.

The influence of the number of training samples 
on the performance of our model is also studied.
For each category, we randomly select training sets of size
5,10,15,20,25 and 30.
\figurename~\ref{fig:15a} shows the performance as
a function of the number of training images.
The results reported are averaged over all categories
with 50 features extracted from each training image.
The performance is compared with other published results.

\begin{figure}
\centering
  \subfloat[Comparison with other methods]{
    \includegraphics[width=0.9\linewidth]{images/fig-15a.pdf}
    \label{fig:15a}}\\
  \subfloat[Comparison with other features]{
    \includegraphics[width=0.9\linewidth]{images/fig-15b.pdf}
    \label{fig:15b}}
\caption{Comparison of performance.}
\label{fig:15}
\end{figure}

\figurename~\ref{fig:15b} shows
the performance of our model over the 101 categories
compared with SIFT features \cite{lowe1999},
PHOW features \cite{lazebnik2006},
and biologically inspired HMAX model \cite{serre2007}.
The results are obtained with 30 training images randomly selected 
from each category.
SIFT features and neural features generated with our model
are passed to an SVM classifier.
The images passed to the HMAX model are resized 
to a uniform height of 140 pixels as recommended by \cite{serre2007}.
Our model exhibits competitive performance 
compared with other models.
As shown in \figurename~\ref{fig:15b},
the SIFT performance points over the 101 categories
are almost completely to the left of the equivalent line.
The PHOW performance points spread through 
a wider range from $0.5$ to $1$ indicating 
the lack of robustness compared with our model.
Compared with the biologically inspired HMAX model,
our performance is also better for more than half
of the 101 categories.
